---
title: "p8105_hw3_zl3543"
author: "Zihan Lin"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggridges)
library(hexbin)
library(patchwork)
library(reshape2)
library(knitr)
```

### Problem 1

```{r}
library(p8105.datasets)
data("ny_noaa")
```

This dataset contains `r nrow(ny_noaa)` rows and `r ncol(ny_noaa)` columns. Variables include weather station id, date of observation,  (tenths of mm), snowfall (mm), snow depth (mm), and min and max temperature (tenths of degrees C).

Below we clean the data, creating separate variables for year, month, and day and converting `tmax` and `tmin` to numeric. We find that 0 is the most commonly observed value for snowfall. This is because most days of the year, it does not snow at all in NY. The second most commonly observed value is `NA`, indicating missingness. Other common values are 13, 25, and 51, suggesting that snowfall is originally recorded in fractions of an inch and converted to mm.

```{r}
# Data cleaning: separate year, month, and day, and convert units
ny_noaa %>% 
  count(snow) %>%
  arrange(desc(n))

ny_noaa_clean <-
  ny_noaa %>% 
  separate(date, into = c("year", "month", "day"), convert = TRUE) %>% 
  mutate(
    tmax = as.numeric(tmax) / 10,   # Convert from tenths of degrees C to degrees C
    tmin = as.numeric(tmin) / 10,   # Convert from tenths of degrees C to degrees C
    prcp = as.numeric(prcp) / 10    # Convert from tenths of mm to mm
  )

# Find the most common snowfall values
snowfall_common <- ny_noaa_clean %>%
  filter(!is.na(snow)) %>%
  count(snow) %>%
  arrange(desc(n)) %>%
  head(5)

print(snowfall_common)

# Plot the most common snowfall values
ggplot(snowfall_common, aes(x = factor(snow), y = n)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Most Common Snowfall Values",
    x = "Snowfall (mm)",
    y = "Frequency"
  ) +
  theme_minimal()
```

Below is a two-panel plot showing the average max temperature in January and in July in each station across years. As expected, the mean temperature in January is much lower than the mean temperature in July for all stations and across all years. All stations appear to follow similar trends of temperature peaks and valleys within a month across the years, i.e. when one station has a high monthly mean temperature for a given year, most other stations also have a high monthly mean temperature for that year. We do see one uncharacteristically cold station in July of 1987 or 1988, as well as a few other less drastic outliers.

```{r}
# Create a plot for average max temperature in January and July
ny_noaa_clean %>% 
  group_by(id, year, month) %>% 
  filter(month %in% c(1, 7)) %>% 
  summarize(mean_tmax = mean(tmax, na.rm = TRUE, color = id)) %>% 
  ggplot(aes(x = year, y = mean_tmax, group = id)) + geom_point() + geom_path() +
  facet_grid(~month) +
  labs(title = "Mean monthly temperature for each station across years for January and July")
```

Below we show a two-panel plot including (i) a hex plot of `tmax` vs `tmin` for the full dataset; and (ii) a ridge plot showing the distribution of snowfall values (in mm) greater than 0 and less than 100 separately by year. 

From the hex plot we see that while there is some variability, the majority of the data cluster tightly in the center of the distribution. In relatively rare cases, it seems that `tmax` is less than `tmin`, which raises questions about data recording and quality.

From the ridge plot, we see a multimodal density of snowfall within a given year. Most stations see between 0 and  35 mm of snow in a year. Then there is a another group of stations that see about 45 mm of snow, and another group that sees nearly 80 mm. It is likely this multimodality stems from the conversion of measurements in one system (fractions of an inch) to another (using the metric system), which was also noted in the table of common values. 

```{r}
# Hexbin plot for tmax vs tmin
ggplot(ny_noaa_clean %>% filter(!is.na(tmax) & !is.na(tmin)), aes(x = tmin, y = tmax)) +
  geom_hex(bins = 50) +
  labs(
    title = "tmax vs tmin",
    x = "Min Temperature (°C)",
    y = "Max Temperature (°C)"
  ) +
  scale_fill_viridis_c() +  
  theme_minimal()

# Ridgeline plot for snowfall distribution by year (0 to 100 mm)
ny_noaa_clean %>%
  filter(snow > 0 & snow <= 100) %>%  # Filter to include snowfall between 0 and 100 mm
  ggplot(aes(x = snow, y = factor(year), fill = after_stat(x))) +  # Use after_stat(x)
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +  # Gradient color
  labs(
    title = "Distribution of Snowfall Over the Years (0 to 100 mm)",
    x = "Snowfall (mm)",
    y = "Year"
  ) +
  scale_fill_viridis_c(option = "C", name = "Snowfall (mm)") +  # Apply viridis color palette
  theme_ridges() +
  theme_minimal() +
  xlim(0, 100) +  # Set x-axis limits to range from 0 to 100 mm
  theme(legend.position = "right")  # Legend positioned to the right
```

### Problem 2

Load, tidy, merge, and otherwise organize the data sets, which includes all originally observed variables; exclude participants less than 21 years of age, and those with missing demographic data; and encode data with reasonable variable classes

```{r}
# Load demographic and accelerometer datasets
covar_data <- read_csv("data/nhanes_covar.csv", skip = 4)
accel_data <- read_csv("data/nhanes_accel.csv")

# Rename the columns for the demographic data for easier access
colnames(covar_data) <- c("SEQN", "sex", "age", "BMI", "education")

# Merge the two datasets using the SEQN column
merged_data <- inner_join(accel_data, covar_data, by = "SEQN")

# Filter participants who are older than 21 and remove rows with missing values
filtered_data <- merged_data %>%
  filter(age >= 21) %>%
  drop_na()

# Check the structure of the filtered data
str(filtered_data)

# Save the cleaned and filtered dataset for further use
write_csv(filtered_data, "cleaned_filtered_data.csv")
```

From the reader-friendly table for the number of men and women in each education category, we found that most people have a "more than high school" education level. There are more female than male in both the category of "less than high school" and the category of "more than high school". For the plot that shows age distributions by sex and education, there are noticeable differences between males and females across all education levels. In the "less than high school" and "high school equivalent" groups, the distribution is more balanced between men and women. However, in the "more than high school" group, women are younger, and men tend to be older. Also, participants with higher education levels (more than high school) tend to be younger, while those with less education are more likely to be older. This suggests a correlation between younger age groups and higher levels of education, which aligns with trends in modern education accessibility and achievement.

```{r}
# Create a summary table for the number of men and women in each education category
summary_table <- filtered_data %>%
  group_by(education, sex) %>%
  summarise(count = n()) %>%
  mutate(sex = ifelse(sex == 1, "Male", "Female")) %>%
  select(education, sex, count)

# Pivot the table to make education levels as columns
summary_table <- summary_table %>%
  pivot_wider(names_from = education, values_from = count, 
              names_glue = "{education}")

# Print the summary table
print(kable(summary_table, caption = "Number of Men and Women in Each Education Category"))

# Recode the education and sex variables for better readability
filtered_data$education <- factor(filtered_data$education, 
                                  levels = c(1, 2, 3),
                                  labels = c("Less than high school", 
                                             "High school equivalent", 
                                             "More than high school"))
filtered_data$sex <- factor(filtered_data$sex, 
                            levels = c(1, 2),
                            labels = c("Male", "Female"))

# Create a violin plot to show age distributions by sex and education
ggplot(filtered_data, aes(x = age, fill = sex)) +
  geom_density(alpha = 0.6) +
  facet_wrap(~education) +
  labs(title = "Distribution of Age by Sex and Education", 
       x = "Age", 
       y = "Density", 
       fill = "Sex") +
  theme_minimal()

```

The plot highlights the interplay between age, sex, and education on physical activity levels. Higher education seems to be associated with higher levels of physical activity across the lifespan, and while both sexes experience a decline in activity with age, women tend to maintain higher activity levels for longer, particularly in higher education groups. 

```{r}
# Aggregate data to get total daily activity for each participant
filtered_data <- filtered_data %>%
  rowwise() %>%
  mutate(total_activity = sum(c_across(starts_with("min"))))  # Sum all minute columns for total activity

# Create the plot
ggplot(filtered_data, aes(x = age, y = total_activity, color = sex)) +
  geom_point(alpha = 0.5) +  # Scatter plot to show individual data points
  geom_smooth(method = "loess", se = FALSE) +  # Add a smooth trend line without confidence interval
  facet_wrap(~education) +  # Separate panels for each education level
  labs(
    title = "Total Daily Activity vs Age by Sex and Education Level",
    x = "Age",
    y = "Total Daily Activity (MIMS)",
    color = "Sex"
  ) +
  theme_minimal() +  # Use a minimal theme for a cleaner look
  theme(
    strip.text = element_text(face = "bold", size = 12),  # Bold facet titles
    plot.title = element_text(face = "bold", size = 14)   # Bold plot title
  )
```

From the 24hr activity plot, we found that the higher the education level, the more structured and prolonged the daily activity appears to be, especially in the afternoon. Participants with higher education maintain higher activity levels for longer periods. In the meantime, women tend to exhibit slightly higher activity levels than men, particularly during the peak periods of the day across all education levels. Despite minor differences in peak times and activity intensities, the general pattern of rising activity in the morning, peaking in the early afternoon, and tapering off into the evening is consistent across all groups.

```{r}
# Gather all minute columns into a long format (minute, activity)
activity_long <- filtered_data %>%
  pivot_longer(cols = starts_with("min"), names_to = "minute", values_to = "activity") %>%
  mutate(minute = as.numeric(gsub("min", "", minute)))  # Convert minute columns from 'min1', 'min2', ... to numbers

# Create the 24-hour activity plot for each education level
ggplot(activity_long, aes(x = minute, y = activity, color = sex)) +
  geom_line(alpha = 0.6, stat = "smooth", method = "loess", se = FALSE) +  # Smooth trends to make patterns clearer
  facet_wrap(~education) +  # Separate panels by education level
  labs(
    title = "24-Hour Activity Time Courses by Sex and Education Level",
    x = "Minute of the Day",
    y = "Activity (MIMS)",
    color = "Sex"
  ) +
  theme_minimal() +  # Use a minimal theme for a cleaner look
  theme(
    strip.text = element_text(face = "bold", size = 12),  # Bold facet titles
    plot.title = element_text(face = "bold", size = 14)   # Bold plot title
  )
```